Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-04-12 pkg/disttree/inst/draft/draft.Rnw
added missing dots at sentence endings
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2148 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a16fac85a966d6a4caf60b2b758df112d1c8a1d8

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{Breiman+Friedman+Stone:1984}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{fig:tree}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{Breiman:2001}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{fig:tree}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from GAMLSS modeling -- as outlined
in Figure~\ref{fig:gam} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional regression trees} (for short: \emph{distributional trees}) split the data into more homogeneous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left)
  \item \emph{Distributional regression forests} (for short: \emph{distributional forests}) utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{fig:tree},
    top right).
\end{itemize}
In the following, particular focus is given to distributional forests as a method for obtaining
probabilistic forecasts by leveraging the strengths of random forests: the ability to
capture both smooth and abruptly changing functions along with simultaneous selection of
variables and possibly complex interactions. Thus, these properties make the method particularly
appealing in case of many covariates with unknown effects and interactions where it would be 
challenging to specify a distributional regression model like GAMLSS.

@@

To sum up, distributional trees are fitted recursively via:
\begin{enumerate}
\item Estimate $\bm{\hat \theta}$ via maximum likelihood for the observations
  in the current subsample.
\item Test for associations or instabilities of the scores $s(\bm{\hat \theta}, y_i)$ 
  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
\item Split the sample along the partitioning variable $Z_l^*$ with the
  strongest association or instability. Choose the breakpoint with the highest
  improvement in the log-likelihood or the highest discrepancy .
\item Repeat steps 1--3 recursively in the subsamples until these become too
  small or there is no significant association/instability (or some other
  stopping criterion is reached).
\end{enumerate}
%
Different inference techniques can be used for assessing the association between
scores and covariates in step~3. In the following we use the general class of
permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} which is
also the basis of conditional inference trees \citep[CTree,][]{Hothorn+Hornik+Zeileis:2006}.
Alternatively, one could use asymptotic M-fluctuation tests for parameter
instability \citep{Zeileis+Hornik:2007} as in model-based recursive partitioning
\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}. More details are provided in
Appendix~\ref{app:tree}.

>>>>

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{Breiman+Friedman+Stone:1984}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{fig:tree}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{Breiman:2001}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{fig:tree}, top middle).
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from GAMLSS modeling -- as outlined
in Figure~\ref{fig:gam} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional regression trees} (for short: \emph{distributional trees}) split the data into more homogeneous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left).
  \item \emph{Distributional regression forests} (for short: \emph{distributional forests}) utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{fig:tree},
    top right).
\end{itemize}
In the following, particular focus is given to distributional forests as a method for obtaining
probabilistic forecasts by leveraging the strengths of random forests: the ability to
capture both smooth and abruptly changing functions along with simultaneous selection of
variables and possibly complex interactions. Thus, these properties make the method particularly
appealing in case of many covariates with unknown effects and interactions where it would be 
challenging to specify a distributional regression model like GAMLSS.

@@

To sum up, distributional trees are fitted recursively via:
\begin{enumerate}
\item Estimate $\bm{\hat \theta}$ via maximum likelihood for the observations
  in the current subsample.
\item Test for associations or instabilities of the scores $s(\bm{\hat \theta}, y_i)$ 
  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
\item Split the sample along the partitioning variable $Z_l^*$ with the
  strongest association or instability. Choose the breakpoint with the highest
  improvement in the log-likelihood or the highest discrepancy.
\item Repeat steps 1--3 recursively in the subsamples until these become too
  small or there is no significant association/instability (or some other
  stopping criterion is reached).
\end{enumerate}
%
Different inference techniques can be used for assessing the association between
scores and covariates in step~3. In the following we use the general class of
permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} which is
also the basis of conditional inference trees \citep[CTree,][]{Hothorn+Hornik+Zeileis:2006}.
Alternatively, one could use asymptotic M-fluctuation tests for parameter
instability \citep{Zeileis+Hornik:2007} as in model-based recursive partitioning
\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}. More details are provided in
Appendix~\ref{app:tree}.
